---
title: "Diagnostic Plots with ggResidpanel"
author: "Katherine Goode and Kathleen Rey"
date: "2018/04/07"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to ggResidpanel

The ggResidpanel package was designed to allow users to easily view multiple diagnostic plots at a time in the ggplot2 framework. The package was motivated by the residual panel plots SAS procedures provide...

# Installing ggResidpanel

In order to install...

# How to Use ggResidpanel

The main function inside of ggResidpanel is `resid_panel`...

# Examples with ggResidpanel

```{r}
# Load the library
library(ggResidpanel)

# First six rows of the mtcars data
head(mtcars)

# Fit a model to predict mpg
mpg_model <- lm(mpg ~ wt + hp, data = mtcars)
```

## SAS Panel Option

```{r}
resid_panel(mpg_model, bins = 25)
```

## R Panel Option

```{r}
resid_panel(mpg_model, bins = 25, plots = "R")
```

## All Plots Panel Option

```{r}
resid_panel(mpg_model, bins = 25, plots = "all")
```

## Individual Plots Panel Option

```{r}
resid_panel(mpg_model, plots = "residplot")
resid_panel(mpg_model, plots = c("residplot", "hist"), bins = 25)
```
